Improved state-space models for inference about spatial and temporal variation in abundance from count data

### 2013-10-14 Dail_Mad_ext.tex
Edits from Andy

\documentclass[12pt]{article}
\renewcommand*\abstractname{Summary}
\usepackage[a4paper,hmargin={2.4cm,2.4cm},vmargin={2.4cm,2.4cm}]{geometry}
\usepackage[mathlines,displaymath]{lineno}
\usepackage{setspace}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{bm}
\usepackage{graphicx}
\usepackage[round,colon,authoryear]{natbib}
\usepackage{color, soul}
\usepackage{fancyvrb}
\usepackage{verbatim}
\usepackage[pdftex,colorlinks=true,citecolor=black,linkcolor=black,urlcolor=black,pdftex,pdfstartview=FitH,bookmarks=true]{hyperref}
\usepackage{rotating}

@@

\author{Jeffrey A. Hostetler\\ %\footnote{Contact:
%     \url{HostetlerJ@si.edu}, +001-202-633-4205}\\
        \small Migratory Bird Center \\
        \small Smithsonian Conservation Biology Institute \\
        \small National Zoological Park, MRC 5503 \\
        \small Washington, DC 20013-7012\\
   \and Richard B. Chandler \footnote{Contact:
     \url{rbchan@uga.edu}, +001-706-542-5815} \\
        \small USGS Patuxent Wildlife Research Center \\
        \small Laurel, MD 20708}
\maketitle

@@

\vspace{0.5cm}

@@

@@

A simple state-space models can be described as follows.
Let $N_t$ be the abundance of a species during year $t$, for
$t=1,\hdots,T$, and let $X_t$ be
the observed data, which differs from $N_t$ due to observation error,
a random effect denoted $\zeta_t$. Temporal variation in $N_t$ is
modeled using a population growth model, $\mu(N_{t-1})$
and random process variation denoted $\grave{o}_t$.
The growth model may be density-dependent, as in the case of the 
Ricker-logistic model, or it might be density-independent, such as when growth
is exponential.
%, in which $\mu(N_{t-1}) = N_{t-1}e^{r}$ where $r$ is the intrinsic rate of increase.
The full model can now be written as:
\begin{subequations}
  \label{eq:ss1}
  \begin{align}
    N_1 &= X_1 \label{eq:ss1a} \\
N_t &= \mu(N_{t-1}) + \grave{o}_{t-1} \quad \text{for} \;
t=2,\hdots,T \label{eq:ss1b} \\
X_t &= N_t + \zeta_t \qquad \qquad \;\, \text{for} \;
t=1,\hdots,T \label{eq:ss1c}
  \end{align}
\end{subequations}
where $\grave{o}_t$ is the random effect allowing for process
variation unaccounted for by the deterministic model. In classical
state-space models, the two sets of random effects
are assumed to be i.i.d. Gaussian deviates:
$\grave{o}_t \sim \mathrm{N}(0, \sigma)$ and
$\zeta_t \sim \mathrm{N}(0, \tau)$. Standard practice is to
ignore process variation associated with $N_1$, as
indicated by Eq~\ref{eq:ss1a}. Alternatively, the population is
assumed to be at equilibrium such that the $N_1$ can be regarded as an
outcome of the equilibrium distribution.

@@

A more serious problem %than the ones associated with the Gaussian assumptions 
is that the parameters of the simple state-space models
such as Eq~\ref{eq:ss1} are not identifiable in many
circumstances \citep{polansky_etal:2009}. Recall that the
Eq.~\ref{eq:ss1} did not specify distributions for $N_1$, % or $X_1$;
which $N_1$ is a random variable. Thus, uncertainty exists
that should be accounted for in the model.  
%are needed to fully specify the model.
%To be more specific, a fully-specified state-space model requires at least
%three probability distributions, which we represent
%using bracket notation (i.e., $[Y|\Omega]$ is interpreted
%as the probability distribution of the random variable $Y$ given the
%parameter $\Omega$). The three probability distributions required for a
%state-space model are:
%%%% JEFF, does this extra sentence help? This notation comes from
%%%% several of the original state-space papers and is widespread in
%%%% books like Andy's and Bill's.
%\hl{[I'm not familiar with bracket notation,
% so these equations are confusing to me. How widely used is this
% notation in ecology? Will others likely understand this? Also, we might want
%to save space by trimming a bunch of this discussion back down.]}
%\begin{subequations}
%  \label{eq:ss2}
%  \begin{align}
%    [N_1&|\bm{\theta}] \label{eq:ss2a} \\ \label{eq:ss2b}
%[N_t&|N_{t-1},\bm{\Theta}] \quad \text{for} \; t=2,\hdots,T \\
%\label{eq:ss2c}
%    [X_t&|N_t,\bm{p}]  \qquad \; \text{for} \; t=1,\hdots,T
%  \end{align}
%\end{subequations}
%where $\bm{\theta}$ are the process variation parameters for the
%initial state, $\bm{\Theta}$ are the process variation parameters
%influencing how abundance changes over time, and $\bm{p}$ are the
%observation error parameters. As mentioned previously, Eq.~\ref{eq:ss2b}
%and Eq.~\ref{eq:ss2c} are assumed to be Gaussian in classical state-space
%models, but what should be the distribution for Eq.~\ref{eq:ss2a}?
%The three lines of general model
%described by Eq.~\ref{eq:ss2} correspond to the three equations shown
%for the specific example in Eq~\ref{eq:ss1}. However, in this case,
%we allow $N_1$ to be a random variable. But what should its
%istribution, $[N_1|\bm{\theta}]$ be?
%And how could $\bm{\theta}$ be estimated given that there is only a single observation
%available? Often, these problems are simply assumed away. For
%example, sometimes researchers assume that there is no process variation in
%the first year, although this seems unjustified. Alternatively, the
%population may be assumed to be at equilibrium such that the expected
%value of $N_t$ is constant through time. Although this 
%makes the parameters $\bm{\theta}$ identifiable, assuming equilibrium
%defeats the objective of many studies of population dynamics, namely
%determining why a population varies over time.

@@

The DM model requires count data collected at $R$ sites, each of
which is surveyed on $T$ primary sampling periods. 
%A site is ideally a
%well-defined region of a study area, such as a wetland or a
%patch of early-successional habitat, although it could be an arbitrarily
%defined region such as a randomly located survey plots. The time
%frame of the study is arbitrary, but in general
%it will be sufficiently long to allow for temporal variation in
%abundance. For instance, a primary sampling period could be a
%one month long breeding season, which could be surveyed once per
%year for $T$ years. If the a site is surveyed on $J$ occasions during a
%primary period, these are called secondary sampling occasions.
%In the case where no secondary sampling was conducted, i.e. $J=1$, 
Let $X_{it}: i=1,\hdots,R; t=1,\hdots,T$ denote the count data
at site $i$ and primary period $t$. 
%If secondary sampling periods were
%used, an extra dimension is added so that we have $X_{ijt}:
%j=1,\hdots,J$. 
Typically, we know $X_{it}$ will be lower than the
actual quantity of interest, abundance,
denoted $N_{it}$. In cases where detection probability is perfect,
abundance is observed directly such that $X_{it} \equiv N_{it}$.

@@

Recall that conventional state-space models assume that the
distribution for the initial time period,
$[N_1 | \bm{\theta}]$, is either the equilibrium
distribution, or has zero variance.
In contrast, \citet{dail_madsen:2011} proposed modeling $N_{i1}$
as either a Poisson or negative binomial random variable:
\begin{gather}
N_{i1} \sim \mathrm{Pois}(\Lambda) \nonumber \\
\text{or} \nonumber \\
N_{i1} \sim \mathrm{NB}(\Lambda, \alpha)
\label{eq:N1}
\end{gather}
where $\Lambda_i$ is the expected abundance at site $i$ during
year 1.
The Poisson distribution assumes that the mean of $N_{i1}$ is
equal to its variance, whereas the negative binomial distribution allows the
variance to be greater than the mean with the amount of
overdispersion determined by the parameter $\alpha$.
%In the Poisson case, the equivalent expression of
%$[N_1 | \bm{\theta}]$ (Eq.~\ref{eq:ss2a}) is
%$[N_{i1} | \Lambda] =\frac{\Lambda^{N_{i1}}e^{-\Lambda}}{N_{i1}!}$.
%\begin{equation}
%[N_{i1} | \Lambda] =
%\frac{\Lambda^{N_{i1}}e^{-\Lambda}}{N_{i1}!}.
%  \label{eq:N1br}
%\end{equation}
% Maybe we don't need the Poisson distriubtion equation? - Jeff

@@

The DM model assumes that abundance in time $t$ is a function of
abundance in time $t-1$, i.e. abundance at each site evolves as a
first order Markovian process. %, although higher order processes are also possible.
\citet{dail_madsen:2011} considered several models to describe the temporal dynamics;
however, in each case they modeled $N_t$ as the sum of two random variables:
$S_{it}$, the number of indivduals surviving from $t-1$ and not
emigrating; and $G_{it}$ the number of new individuals entering
the population. Their most general model was
\begin{equation}
\left.\begin{aligned}
S_{it}|N_{it-1} &\sim \mathrm{Bin}(N_{it-1}, \omega) \\
G_{it}|N_{it-1} &\sim \mathrm{Pois}(\gamma(N_{it-1})) \\
N_{it} &= S_{it}+G_{it}
\end{aligned}\right\} \quad \text{for} \; t=2,\hdots,T
\label{eq:Nt}
\end{equation}
where $\omega$ is the apparent survival probability and $\gamma$
is the recruitment rate (which can depend on $N_{it-1}$).
\citet{dail_madsen:2011} proposed three
models for $\gamma$: the constant model,
$G_{it} \sim \mathrm{Pois}(\gamma)$ where recruitment does not
depend on $N_{it-1}$, and which simulates a ``propagule rain'' of new
individuals; the autoregressive model, $G_{it} \sim
\mathrm{Pois}(\gamma(N_{it-1}))$, which
simulates geometric or density independent growth; and the
``no-trend'' model, $\gamma = (1-\omega)\Lambda$, which keeps
expected abundance constant over time. Covariates of
$\omega$ and $\gamma$ can be easily accommodated, for example
using logit- and log-linear models respectively.

@@

%Eqs.~\ref{eq:N1} and \ref{eq:Nt} fully specify the state model,
%i.e. the model for abundance dynamics. This state
%model has allowed us to replace the initial abundance
%distribution and transition models of conventional state-space models, with
%alternatives that (1) allow for modeling spatial variation in
%abundance, (2) respect the discrete nature of the counts, and
%(3) have parameters that are estimable.
The one
remaining task is to replace the observation model
$X_t = N_t + \zeta_t$ (Eq.~\ref{eq:ss1c}) with one that can be
more easily interpreted. The observation
model assumes that individuals are missed due to imperfect
detection. The simplest model for imperfect detection is
\begin{equation}
  X_{it} \sim \mathrm{Bin}(N_{it}, p)
  \label{eq:p1}
\end{equation}
%or equivalently,
%\begin{equation}
%[X_{it} | N_{it}, \bm{p}] = \binom{N_{it}}{X_{it}}
%\bm{p}^{X_{it}}(1-\bm{p})^{N_{it}-X_{it}}.
%  \label{eq:1}
%\end{equation}
Under this model, $\bm{p}$ is interpreted as the detection
probability parameter.

@@

Partitioning population growth into survival and recruitment
is ideal in terms of providing a mechanistic description of population
dynamics, but this is not always possible using simple count data,
especially when the sites are not closed with respect to immigration
and emigration. %Furthermore, a
%simpler model would have several merits: faster running time, fewer
%total model combinations, and possibly more realistic estimates.
If the mechanistic model must be sacrificed, we suggest replacing it
with classical population growth models that have a long history in
ecology. The exponential growth model is the simplest: 
$f(N_{it-1}) = N_{it-1}e^r$ where $r$ is the intrinsic
rate of increase. %Although demographic and environmental stochasticity
%in $r$ could be modeled as functions of covariates or random effects, a
%simpler way of 
To allow for demographic stochasticity, we can regard $N_{it}$ as a Poisson 
random variable, simplifying Eq.~\ref{eq:Nt} to:
\begin{equation}
  N_{it} \sim \text{Pois}(\exp(r)N_{it-1}).
\label{eq:exp}
\end{equation}
This,
like the geometric-recruitment version of the model, is a variant on a simple
density-independent, exponential model of population growth.
Density-dependent versions of the model are also possible.  For
example:
\begin{equation}
  N_{it} \sim \text{Pois}(N_{it-1}\exp(r(1-N_{it-1}/K)))
\label{eq:rick}
\end{equation}
where $K$ is the stable equilibrium of the population and $r$ is
the
instantaneous population growth rate at low population
densities, and
both parameters are constrained to be positive. This is a
stochastic version
of the Ricker-logistic population growth model
\citep{ricker:1954}. We also
implemented a modified Gompertz-logistic density-dependent
model \citep{hart_gotelli:2011}:
\begin{equation}
N_{it} \sim
\text{Pois}(N_{it-1}\exp(r(1-\log(N_{it-1}+1)/\log(K+1))))
\label{eq:gomp}
\end{equation}
Here the interpretations of $r$ and $K$ are similar to in the
Ricker-logistic model.
Because a single Poisson distribution controls the
distribution of $N_{it}$ in each of these models, the discrete
convolution used by \citet{dail_madsen:2011} is not required,
speeding up processing time.

@@

The geometric-recruitment, exponential, Ricker-logistic, and Gompertz-logistic
versions of the DM models all share a common feature (or bug):
once the population at a site reaches 0, it must remain at 0
because all contributions to population growth are assumed to be
local.  We therefore generalized these models to
include both internal and external (immigration) contributions
to population growth.  The exponential plus immigration
model is:
\begin{equation}
  N_{it} \sim \text{Poisson}(\exp(r)N_{it-1} + \iota)
  \label{eq:expimm2}
\end{equation}
where $\iota$ represents the immigration rate, and is
constrained to be
positive (this is equivalent to separate Poisson processes for
growth and immigration).  %This model is close to
%the constant-recruitment DM model (Eq.~\ref{eq:Nt}), with $\exp(r)$ instead
%of $\omega$ and $\iota$ instead of $\gamma$, except that the first
%process is Poisson distributed instead of binomial. 
The geometric-recruitment, Ricker-logistic,
and Gompertz-logistic models can be extended to allow for immigration in the
same way.   Deterministic versions of the Ricker-logistic + immigration and the 
Gompertz-logistic + immigration do reach stable equilibriums, but the equilibriums 
are not at $K$ and cannot be solved for analytically \citep{otto_day:2007}.  Therefore, we
refer to $K$ in these models as the semi-equilibrium abundance.
\begin{comment}
Similarly, the Ricker-logistic + immigration model can be represented as:
the
Gompertz-logistic + immigration model as: (13) and the geometric-recruitment +  immigration model as: (14)
\end{comment}

@@

The zero-inflated Poisson distribution can be applied to recruitment and 
population growth terms as well as initial abundance. For example, 
the recruitment term of the constant-recruitment DM model
(Eq.~\ref{eq:Nt}) can be modified as follows:
\begin{equation}
G_{it} \sim \left\{
\begin{aligned}
0 &\; \text{with probability} \; \psi \\
\mathrm{Poisson}(\gamma) &\; \text{with probability} \; (1-\psi)\end{aligned} \right.
\label{eq:ZIPts}
\end{equation}
\begin{comment}
When both initial abundance and dynamics are modeled as
zero-inflated,
$\psi$ can be modeled as time-varying or not (in either case,
$\psi$ can also be
modeled as varying over space). In the latter case, it makes
sense
for the model to "remember" which sites are inside the range of
the
species between the initial abundance stage and each time step
of the
dynamics stage (see code sample?).  Adding zero-inflation to the
dynamics of the geometric-recruitment, exponential, Ricker-logistic, or Gompertz-logistic
models
(without immigration) should have no effect when initial
abundance is
also zero-inflated unless $\psi$ is time-varying (because these
dynamic
models cannot recover from zero abundance).
\end{comment}
We have implemented
zero-inflated dynamics in the Bayesian framework using program
\textbf{JAGS} \citep[version 3.2.0]{plummer:2003}.
%(version 3.2.0; Plummer 2003)

\subsection{Environmental and demographic stochasticity}
\begin{comment}
Even though the DM model is conceptually simple, even in
its most stripped down form it includes many random effects (the $N$'s)
making it challenging to compute the likelihood or implement MCMC
algorithms. Nonetheless, in some cases additional random effects
may be of interest. For example, in conventional state-space models,
environmental stochasticity is often modeled as $r_t \sim
\text{Norm}(\mu_r, \sigma_r)$.
\end{comment}
The original DM models include demographic stochasticity --- variation in population growth 
rate due to the randomness of birth and death processes --- through random 
binomial draws for survival and Poisson draws for recruitment.  Our population
growth extensions use a single Poisson draw for abundance: $N_{it} \sim 
\text{Pois}(f(N_{it-1}))$.  This is also a commonly used (and mathematically equivalent) 
method of modeling demographic stochasticity 
\citep{bonsall_hastings:2004,melbourne_hastings:2008}.

One way to model environmental stochasticity --- variation in population growth
rate due to stochastic environmental conditions --- is by explicitly
modeling dynamic parameters as functions of environmental covariates, such
as weather.  For example, in the Gompertz-logistic model (Eq.~\ref{eq:gomp}), $r$ could be a 
function of temperature ($T$) and $K$ a function of rainfall ($R$):
\begin{gather}
\log(r_{it}) = \beta^{r}_{0} + \beta^{r}_{1}T_{it} \nonumber \\
\log(K_{it}) = \beta^{K}_{0} + \beta^{K}_{1}R_{it}  
%N_{it} \sim
%\text{Pois}(N_{it-1}\exp(r_{it-1}(1-\log(N_{it-1}+1)/\log(K_{it-1}+1)))) \nonumber
\label{eq:weather}
\end{gather}

>>>>

\documentclass[12pt]{article}
\renewcommand*\abstractname{Summary}
\usepackage[a4paper,hmargin={2.4cm,2.4cm},vmargin={2.4cm,2.4cm}]{geometry}
\usepackage[mathlines]{lineno} %,displaymath
\usepackage{setspace}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{bm}
\usepackage{graphicx}
\usepackage[round,colon,authoryear]{natbib}
\usepackage{color, soul}
\usepackage{fancyvrb}
\usepackage{verbatim}
\usepackage[pdftex,colorlinks=true,citecolor=black,linkcolor=black,urlcolor=black,pdftex,pdfstartview=FitH,bookmarks=true]{hyperref}
\usepackage{rotating}

@@

%\doublespacing
\author{Jeffrey A. Hostetler\\ %\footnote{Contact:
%     \url{HostetlerJ@si.edu}, +001-202-633-4205}\\
        \small Migratory Bird Center \\
        \small Smithsonian Conservation Biology Institute \\
        \small National Zoological Park, MRC 5503 \\
        \small Washington, DC 20013-7012\\
   \and Richard B. Chandler \footnote{Contact:
     \url{rbchan@uga.edu}, +001-706-542-5815} \\
        \small USGS Patuxent Wildlife Research Center \\
        \small Laurel, MD 20708}
\maketitle

@@

%\vspace{0.5cm}

@@

%\newpage
\vspace{0.5cm}

@@

A simple state-space models can be described as follows.
Let $N_t$ be the abundance of a species during year $t$, for
$t=1,\hdots,T$, and let $X_t$ be
the observed data, which differs from $N_t$ due to observation error,
a random effect denoted $\zeta_t$. Temporal variation in $N_t$ is
modeled using a population growth model, $\mu(N_{t-1})$
and random process variation denoted $\eta_t$.
% Richard, is switching this from $\grave{o}$ to $\eta$ okay?  Andy thought
% the o with accent was weird and I think I agree.  $\nu$ seems standard,
% but we use it for a slightly different meaning (ES only) later.  
The growth model may be density-dependent, as in the case of the 
Ricker-logistic model, or it might be density-independent, such as when growth
is exponential.
%, in which $\mu(N_{t-1}) = N_{t-1}e^{r}$ where $r$ is the intrinsic rate of increase.
The full model can now be written as:
\begin{subequations}
  \label{eq:ss1}
  \begin{align}
    N_1 &= X_1 \label{eq:ss1a} \\
N_t &= \mu(N_{t-1}) + \eta_{t-1} \quad \text{for} \;
t=2,\hdots,T \label{eq:ss1b} \\
X_t &= N_t + \zeta_t \qquad \qquad \;\, \text{for} \;
t=1,\hdots,T \label{eq:ss1c}
  \end{align}
\end{subequations}
where $\eta_t$ is the random effect allowing for process
variation unaccounted for by the deterministic model. In classical
state-space models, the two sets of random effects
are assumed to be i.i.d. Gaussian deviates:
$\eta_t \sim \mathrm{N}(0, \sigma)$ and
$\zeta_t \sim \mathrm{N}(0, \tau)$. Standard practice is to
ignore process variation associated with $N_1$, as
indicated by Eq~\ref{eq:ss1a}. Alternatively, the population is
assumed to be at equilibrium such that the $N_1$ can be regarded as an
outcome of the equilibrium distribution.

@@

A more serious problem %than the ones associated with the Gaussian assumptions 
is that the parameters of the simple state-space models
such as Eq~\ref{eq:ss1} are not identifiable in many
circumstances \citep{polansky_etal:2009}. Recall that the
Eq.~\ref{eq:ss1} did not specify distributions for $N_1$, % or $X_1$;
although $N_1$ is a random variable. Thus, uncertainty exists
that should be accounted for in the model.  
%are needed to fully specify the model.
To be more specific, a fully-specified state-space model requires at least
three probability distributions, which we represent
using bracket notation (i.e., $[Y|\Omega]$ is interpreted
as the probability distribution of the random variable $Y$ given the
parameter $\Omega$). The three probability distributions required for a
state-space model are:
%%%% JEFF, does this extra sentence help? This notation comes from
%%%% several of the original state-space papers and is widespread in
%%%% books like Andy's and Bill's.
%\hl{[I'm not familiar with bracket notation,
% so these equations are confusing to me. How widely used is this
% notation in ecology? Will others likely understand this? Also, we might want
%to save space by trimming a bunch of this discussion back down.]}
\begin{subequations}
  \label{eq:ss2}
  \begin{align}
    [N_1&|\bm{\theta}] \label{eq:ss2a} \\ \label{eq:ss2b}
[N_t&|N_{t-1},\bm{\Theta}] \quad \text{for} \; t=2,\hdots,T \\
\label{eq:ss2c}
    [X_t&|N_t,\bm{p}]  \qquad \; \text{for} \; t=1,\hdots,T
  \end{align}
\end{subequations}
where $\bm{\theta}$ are the process variation parameters for the
initial state, $\bm{\Theta}$ are the process variation parameters
influencing how abundance changes over time, and $\bm{p}$ are the
observation error parameters. As mentioned previously, Eq.~\ref{eq:ss2b}
and Eq.~\ref{eq:ss2c} are assumed to be Gaussian in classical state-space
models, but what should be the distribution for Eq.~\ref{eq:ss2a}?
%The three lines of general model
%described by Eq.~\ref{eq:ss2} correspond to the three equations shown
%for the specific example in Eq~\ref{eq:ss1}. However, in this case,
%we allow $N_1$ to be a random variable. But what should its
%distribution, $[N_1|\bm{\theta}]$ be?
And how can $\bm{\theta}$ be estimated given that there is only a single observation
available? 
%Often, these problems are simply assumed away. For
%example, sometimes researchers assume that there is no process variation in
%the first year, although this seems unjustified. Alternatively, the
%population may be assumed to be at equilibrium such that the expected
%value of $N_t$ is constant through time. Although this 
%makes the parameters $\bm{\theta}$ identifiable, assuming equilibrium
%defeats the objective of many studies of population dynamics, namely
%determining why a population varies over time.
%% OR
%Sometimes researchers assume that there is no process variation in 
%the first year or that the population is at equilibrium such that the expected
%value of $N_t$ is constant through time, but these approaches are unsatisfactory.

@@

The DM model requires count data collected at $R$ sites, each of
which is surveyed on $T$ primary sampling periods. 
%A site is ideally a
%well-defined region of a study area, such as a wetland or a
%patch of early-successional habitat, although it could be an arbitrarily
%defined region such as a randomly located survey plots. The time
%frame of the study is arbitrary, but in general
%it will be sufficiently long to allow for temporal variation in
%abundance. For instance, a primary sampling period could be a
%one month long breeding season, which could be surveyed once per
%year for $T$ years. If the a site is surveyed on $J$ occasions during a
%primary period, these are called secondary sampling occasions.
%In the case where no secondary sampling was conducted, i.e. $J=1$, 
Let $X_{i,t}: i=1,\hdots,R; t=1,\hdots,T$ denote the count data
at site $i$ and primary period $t$. 
%If secondary sampling periods were
%used, an extra dimension is added so that we have $X_{ijt}:
%j=1,\hdots,J$. 
Typically, we know $X_{i,t}$ will be lower than the
actual quantity of interest, abundance,
denoted $N_{i,t}$. In cases where detection probability is perfect,
abundance is observed directly such that $X_{i,t} \equiv N_{i,t}$.

@@

Recall that conventional state-space models assume that the
distribution for the initial time period,
$[N_1 | \bm{\theta}]$, is either the equilibrium
distribution, or has zero variance.
In contrast, \citet{dail_madsen:2011} proposed modeling $N_{i1}$
as either a Poisson or negative binomial random variable:
\begin{gather}
N_{i,1} \sim \mathrm{Pois}(\Lambda) \nonumber \\
\text{or} \nonumber \\
N_{i,1} \sim \mathrm{NB}(\Lambda, \alpha)
\label{eq:N1}
\end{gather}
where $\Lambda_i$ is the expected abundance at site $i$ during
year 1.
The Poisson distribution assumes that the mean of $N_{i,1}$ is
equal to its variance, whereas the negative binomial distribution allows the
variance to be greater than the mean with the amount of
overdispersion determined by the parameter $\alpha$.
%In the Poisson case, the equivalent expression of
%$[N_1 | \bm{\theta}]$ (Eq.~\ref{eq:ss2a}) is
%$[N_{i1} | \Lambda] =\frac{\Lambda^{N_{i1}}e^{-\Lambda}}{N_{i1}!}$.
%\begin{equation}
%[N_{i1} | \Lambda] =
%\frac{\Lambda^{N_{i1}}e^{-\Lambda}}{N_{i1}!}.
%  \label{eq:N1br}
%\end{equation}
% Maybe we don't need the Poisson distriubtion equation? - Jeff

@@

The DM model assumes that abundance in time $t$ is a function of
abundance in time $t-1$, i.e. abundance at each site evolves as a
first order Markovian process. %, although higher order processes are also possible.
\citet{dail_madsen:2011} considered several models to describe the temporal dynamics;
however, in each case they modeled $N_t$ as the sum of two random variables:
$S_{i,t}$, the number of indivduals surviving from $t-1$ and not
emigrating; and $G_{i,t}$ the number of new individuals entering
the population. Their most general model was
\begin{equation}
\left.\begin{aligned}
S_{i,t}|N_{i,t-1} &\sim \mathrm{Bin}(N_{i,t-1}, \omega) \\
G_{i,t}|N_{i,t-1} &\sim \mathrm{Pois}(\gamma(N_{i,t-1})) \\
N_{i,t} &= S_{i,t}+G_{i,t}
\end{aligned}\right\} \quad \text{for} \; t=2,\hdots,T
\label{eq:Nt}
\end{equation}
where $\omega$ is the apparent survival probability and $\gamma$
is the recruitment rate (which can depend on $N_{i,t-1}$).
\citet{dail_madsen:2011} proposed three
models for $\gamma$: the constant model,
$G_{i,t} \sim \mathrm{Pois}(\gamma)$ where recruitment does not
depend on $N_{i,t-1}$, and which simulates a ``propagule rain'' of new
individuals; the autoregressive model, $G_{i,t} \sim
\mathrm{Pois}(\gamma(N_{i,t-1}))$, which
simulates geometric or density independent growth; and the
``no-trend'' model, $\gamma = (1-\omega)\Lambda$, which keeps
expected abundance constant over time. Covariates of
$\omega$ and $\gamma$ can be easily accommodated, for example
using logit- and log-linear models respectively.

@@

%Eqs.~\ref{eq:N1} and \ref{eq:Nt} fully specify the state model,
%i.e. the model for abundance dynamics. This state
%model has allowed us to replace the initial abundance
%distribution and transition models of conventional state-space models, with
%alternatives that (1) allow for modeling spatial variation in
%abundance, (2) respect the discrete nature of the counts, and
%(3) have parameters that are estimable.
The one
remaining task is to replace the observation model
$X_t = N_t + \zeta_t$ (Eq.~\ref{eq:ss1c}) with one that can be
more easily interpreted. The observation
model assumes that individuals are missed due to imperfect
detection. The simplest model for imperfect detection is
\begin{equation}
  X_{i,t} \sim \mathrm{Bin}(N_{i,t}, p)
  \label{eq:p1}
\end{equation}
%or equivalently,
%\begin{equation}
%[X_{it} | N_{it}, \bm{p}] = \binom{N_{it}}{X_{it}}
%\bm{p}^{X_{it}}(1-\bm{p})^{N_{it}-X_{it}}.
%  \label{eq:1}
%\end{equation}
Under this model, $\bm{p}$ is interpreted as the detection
probability parameter.

@@

Partitioning population growth into survival and recruitment
is ideal in terms of providing a mechanistic description of population
dynamics, but this is not always possible using simple count data,
especially when the sites are not closed with respect to immigration
and emigration. %Furthermore, a
%simpler model would have several merits: faster running time, fewer
%total model combinations, and possibly more realistic estimates.
If the mechanistic model must be sacrificed, we suggest replacing it
with classical population growth models that have a long history in
ecology. The exponential growth model is the simplest: 
$f(N_{i,t-1}) = N_{i,t-1}e^r$ where $r$ is the intrinsic
rate of increase. %Although demographic and environmental stochasticity
%in $r$ could be modeled as functions of covariates or random effects, a
%simpler way of 
To allow for demographic stochasticity, we can regard $N_{i,t}$ as a Poisson 
random variable, simplifying Eq.~\ref{eq:Nt} to:
\begin{equation}
  N_{i,t} \sim \mathrm{Pois}(\exp(r)N_{i,t-1}).
\label{eq:exp}
\end{equation}
This,
like the geometric-recruitment version of the model, is a variant on a simple
density-independent, exponential model of population growth.
Density-dependent versions of the model are also possible.  For
example:
\begin{equation}
  N_{i,t} \sim \mathrm{Pois}(N_{i,t-1}\exp(r(1-N_{i,t-1}/K)))
\label{eq:rick}
\end{equation}
where $K$ is the stable equilibrium of the population and $r$ is
the
instantaneous population growth rate at low population
densities, and
both parameters are constrained to be positive. This is a
stochastic version
of the Ricker-logistic population growth model
\citep{ricker:1954}. We also
implemented a modified Gompertz-logistic density-dependent
model \citep{hart_gotelli:2011}:
\begin{equation}
N_{i,t} \sim
\mathrm{Pois}(N_{i,t-1}\exp(r(1-\log(N_{i,t-1}+1)/\log(K+1))))
\label{eq:gomp}
\end{equation}
Here the interpretations of $r$ and $K$ are similar to in the
Ricker-logistic model.
Because a single Poisson distribution controls the
distribution of $N_{i,t}$ in each of these models, the discrete
convolution used by \citet{dail_madsen:2011} to construct the likelihood is not required,
speeding up processing time.

@@

The geometric-recruitment, exponential, Ricker-logistic, and Gompertz-logistic
versions of the DM models all share a common feature (or bug):
once the population at a site reaches 0, it must remain at 0
because all contributions to population growth are assumed to be
local.  We therefore generalized these models to
include both internal and external (immigration) contributions
to population growth.  The exponential plus immigration
model is:
\begin{equation}
  N_{i,t} \sim \mathrm{Pois}(\exp(r)N_{i,t-1} + \iota)
  \label{eq:expimm2}
\end{equation}
where $\iota$ represents the average number of immigrants per year, and is
constrained to be
positive (this is equivalent to separate Poisson processes for
growth and immigration).  %This model is close to
%the constant-recruitment DM model (Eq.~\ref{eq:Nt}), with $\exp(r)$ instead
%of $\omega$ and $\iota$ instead of $\gamma$, except that the first
%process is Poisson distributed instead of binomial. 
The geometric-recruitment, Ricker-logistic,
and Gompertz-logistic models can be extended to allow for immigration in the
same way.   Deterministic versions of the Ricker-logistic + immigration and the 
Gompertz-logistic + immigration do reach stable equilibriums, but the equilibriums 
are not at $K$ and cannot be solved for analytically \citep{otto_day:2007}.  Therefore, we
refer to $K$ in these models as the semi-equilibrium abundance.
\begin{comment}
Similarly, the Ricker-logistic + immigration model can be represented as:
the
Gompertz-logistic + immigration model as: (13) and the geometric-recruitment +  immigration model as: (14)
\end{comment}

@@

The zero-inflated Poisson distribution can be applied to recruitment and 
population growth terms as well as initial abundance. For example, 
the recruitment term of the constant-recruitment DM model
(Eq.~\ref{eq:Nt}) can be modified as follows:
\begin{equation}
G_{i,t} \sim \left\{
\begin{aligned}
0 &\; \text{with probability} \; \psi \\
\mathrm{Pois}(\gamma) &\; \text{with probability} \; (1-\psi)\end{aligned} \right.
\label{eq:ZIPts}
\end{equation}
\begin{comment}
When both initial abundance and dynamics are modeled as
zero-inflated,
$\psi$ can be modeled as time-varying or not (in either case,
$\psi$ can also be
modeled as varying over space). In the latter case, it makes
sense
for the model to "remember" which sites are inside the range of
the
species between the initial abundance stage and each time step
of the
dynamics stage (see code sample?).  Adding zero-inflation to the
dynamics of the geometric-recruitment, exponential, Ricker-logistic, or Gompertz-logistic
models
(without immigration) should have no effect when initial
abundance is
also zero-inflated unless $\psi$ is time-varying (because these
dynamic
models cannot recover from zero abundance).
\end{comment}
We have implemented
zero-inflated dynamics in the Bayesian framework using program
\textbf{JAGS} \citep[version 3.2.0]{plummer:2003}.
%(version 3.2.0; Plummer 2003)

\subsection{Environmental and demographic stochasticity}
\begin{comment}
Even though the DM model is conceptually simple, even in
its most stripped down form it includes many random effects (the $N$'s)
making it challenging to compute the likelihood or implement MCMC
algorithms. Nonetheless, in some cases additional random effects
may be of interest. For example, in conventional state-space models,
environmental stochasticity is often modeled as $r_t \sim
\mathrm{Norm}(\mu_r, \sigma_r)$.
\end{comment}
The original DM models include demographic stochasticity --- variation in population growth 
rate due to the randomness of birth and death processes --- through random 
binomial draws for survival and Poisson draws for recruitment.  Our population
growth extensions use a single Poisson draw for abundance: $N_{i,t} \sim 
\mathrm{Pois}(f(N_{i,t-1}))$.  This is also a commonly used (and mathematically equivalent) 
method of modeling demographic stochasticity 
\citep{bonsall_hastings:2004,melbourne_hastings:2008}.

One way to model environmental stochasticity --- variation in population growth
rate due to stochastic environmental conditions --- is by explicitly
modeling dynamic parameters as functions of environmental covariates, such
as weather.  For example, in the Gompertz-logistic model (Eq.~\ref{eq:gomp}), $r$ could be a 
function of temperature ($T$) and $K$ a function of rainfall ($R$):
\begin{gather}
\log(r_{i,t}) = \beta^{r}_{0} + \beta^{r}_{1}T_{i,t} \nonumber \\
\log(K_{i,t}) = \beta^{K}_{0} + \beta^{K}_{1}R_{i,t}  
%N_{it} \sim
%\text{Pois}(N_{it-1}\exp(r_{it-1}(1-\log(N_{it-1}+1)/\log(K_{it-1}+1)))) \nonumber
\label{eq:weather}
\end{gather}
